Mixture Models for Single-Cell Assays with Applications to Vaccine Studies

### 2013-06-04 MIMOSA.tex
shorten again

 \author{GREG FINAK$^{1}$, 
 ANDREW MCDAVID$^{1}$,
 PRATIP CHATTOPADHYAY$^{3}$,
 MARIA DOMINGUEZ$^{3}$,
 STEVE DE ROSA$^{1,2}$,
 MARIO ROEDERER$^{3}$,\\
 RAPHAEL GOTTARDO$^{1\ast}$
 \\[4pt]
\textit{ $^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
 $^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
 $^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
 \\[2pt]
 {gfinak@fhcrc.org}}

@@

%Another motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
A motivating example from vaccine research is the need to assess the immunogenicity of a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
T-cells that recognize the antigen become \emph{activated}, produce cytokines that potentiate the immune response, and proliferate and persist in the immune system providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}. This vaccine memory can be interrogated through the intracellular cytokine staining (ICS) assay, by measuring cytokine production in antigen-specific T-cells in response to antigen stimulation~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
 %While not all T-cells can recognize all antigens, those that recognize antigens in the vaccine become \emph{activated} and produce cytokines that potentiate the immune response.
 %After activation, this antigen-specific subpopulation proliferates and can persist in the immune system for some time providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}.

The antigen-specific subpopulations constitute a small fraction of the total number of CD4 and CD8 T-cells (as low as 0.01\%--0.1\%), 
% The intracellular cytokine staining (ICS) assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring intracellular cytokine production in response to cell activation following stimulation by an antigen that closely matches what was present in the original vaccine.
%Individual cells are labelled using fluorescently conjugated antibodies against cell-surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
and a large number of cells must be collected (on the order of 50,000 to 100,000 T-cells) to ensure that changes in these cell populations can be reliably detected.
Then, each individual cell is classified as either positive or negative for each cytokine based on fixed thresholds, and the number of cells matching each phenotype is counted.

@@

\noindent\textbf{MCMC algorithm:} We generated realizations from the posterior distribution via MCMC algorithms~\citep{Walter1996}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs sampling.
Details about the algorithms and implementation for the data presented here are given in Appendix B of the Supplementary Material. 
%We used the
%method of \cite{Raftery:1992vp} and Raftery \citep{Walter1996} to determine
%the number of iterations, based on a short pilot run of the
%sampler. For each dataset presented here, we calculated that no more
%than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. To leave some margin, we used 200,000 iterations after 50,000 burn-in iterations for each dataset explored here.

@@

MIMOSA is fit to dichotomized data  (cells are either positive or negative). While this is the standard approach for ICS data analysis, it is believed that the magnitude (fluorescence intensity) of the signal also carries information. For single-cell gene expression data, most methods have focused on differences in the continuous part of the signal while using the discrete on/off nature for pre-filtering~\citep{Flatz:2011jb}. 
The ability of MIMOSA to identify stimulation-specific expression patterns in the Fluidigm data demonstrates not only the broader utility of the method, but demonstrates that biologically relevant signal is present in the frequency of gene expression (Figure~\ref{fig:fluidigm}). 
We have shown elsewhere that changes in both the frequency of expression and expression level are informative~\citep{McDavid2012}, and extensions of the model that incorporate the continuous component (MFI or gene expression) are warranted in the future. 
Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative subsets.

@@

When changes are two--sided (as with Fluidigm), differences in poly-functional cell populations are not always detectable  since differences may cancel in the marginal populations (Figure~\ref{fig:polyfunctionality} B).
In this case, the use of multivariate models, as our Dirichlet-multinomial model, will be important (Figure~\ref{fig:polyfunctionality} A).
Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
In the case of more than two biomarkers, the number of parameters for our model is $2^{K+1}+1$, which is large even for moderate values of K. Even for  $K=4$, we would need a large number of subjects and cells to properly estimate the 33 parameters. A solution would be to explore alternative model parameterizations that could be used to reduce the number of required parameters.
For example, one could assume that the hyper-parameters are constant across biomarker combinations, \textit{i.e.}, $\alpha^{\{s,u\}}_{k}=\alpha^{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
In reality, such a model would be unrealistic given that certain stimulations are known to induce expression of certain biomarkers more than others.
More exploratory work will need to be done in this area once high dimensional single-cell level data with large number of samples become available.

>>>>

 \author{GREG FINAK$^{1}$, 
 ANDREW MCDAVID$^{1}$,
 PRATIP CHATTOPADHYAY$^{3}$,
 MARIA DOMINGUEZ$^{3}$,
 STEVE DE ROSA$^{1,2}$,
 MARIO ROEDERER$^{3}$,
 RAPHAEL GOTTARDO$^{1\ast}$
 \\[4pt]
\textit{ $^{1}$Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
 $^{2}$HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA\\
 $^{3}$Vaccine Research Center, NIAID, NIH, 40 Convent Drive, Rm 5509, Bethesda, MD 20892}
 \\[2pt]
 {rgottard@fhcrc.org}}

@@

%Another motivating example from vaccine research is the flow cytometric intracellular cytokine staining (ICS) assay, which is used to identify and quantify subjects' immune responses to a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
A motivating example from vaccine research is the need to assess the immunogenicity of a vaccine. Upon vaccination, antigen in the vaccine is taken up and presented to CD4 or CD8 T-cells via antigen presenting cells.
T-cells that recognize the antigen become \emph{activated}, produce cytokines that potentiate the immune response, and proliferate and persist in the immune system providing \emph{memory} that can more rapidly recognize the same antigen in the future~\citep{McKinstry:2010ei}. This vaccine memory can be interrogated through the intracellular cytokine staining (ICS) assay, by measuring antigen-specific cytokine production in response to stimulation~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}. The antigen-specific subpopulations constitute a small fraction of the total number of CD4 and CD8 T-cells (as low as 0.01\%--0.1\%), 
% The intracellular cytokine staining (ICS) assay measures the number of antigen-specific T-cells in PBMC or whole blood by measuring intracellular cytokine production in response to cell activation following stimulation by an antigen that closely matches what was present in the original vaccine.
%Individual cells are labelled using fluorescently conjugated antibodies against cell-surface biomarkers (CD3, CD4, and CD8), used to subset T--cells, and functional biomarkers (cytokines) used to define antigen specific T--cells~\citep{Horton:2007tsa,DeRosa:2004wp,Betts:2006dw}.
and a large number of cells must be collected (on the order of 50,000 to 100,000 T-cells) to ensure that changes in these cell populations can be reliably detected.
Then, each individual cell is classified as either positive or negative for each cytokine based on fixed thresholds, and the number of cells matching each phenotype is counted.

@@

\noindent\textbf{MCMC algorithm:} We generated realizations from the posterior distribution via MCMC algorithms \citep{Gelfand1990}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were performed via Gibbs sampling.
Details about the algorithms and implementation are given in Appendix B of the Supplementary Material. 
%We used the
%method of \cite{Raftery:1992vp} and Raftery \citep{Walter1996} to determine
%the number of iterations, based on a short pilot run of the
%sampler. For each dataset presented here, we calculated that no more
%than about 100,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. To leave some margin, we used 200,000 iterations after 50,000 burn-in iterations for each dataset explored here.

@@

MIMOSA is fit to dichotomized data (cells are positive or negative). While this is the standard approach for ICS data analysis, it is believed that the magnitude (fluorescence intensity) of the signal also carries information. For single-cell gene expression data, most methods have focused on differences in the continuous part of the signal while using the discrete on/off nature for pre-filtering~\citep{Flatz:2011jb}. 
The ability of MIMOSA to identify stimulation-specific expression patterns in the Fluidigm data demonstrates not only the broader utility of the method, but demonstrates that biologically relevant signal is present in the frequency of gene expression (Figure~\ref{fig:fluidigm}). 
We have shown elsewhere that changes in both the frequency of expression and expression level are informative~\citep{McDavid2012}, and extensions of the model that incorporate the continuous component are warranted in the future. 
Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative subsets.

@@

When changes are two-sided (as with Fluidigm), differences in poly-functional cell populations are not always detectable  since differences may cancel in the marginal populations (Figure~\ref{fig:polyfunctionality} B).
In this case, the use of multivariate models, as our Dirichlet-multinomial model, will be important (Figure~\ref{fig:polyfunctionality} A).
Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
In the case of more than two biomarkers, the number of parameters for our model is $2^{K+1}+1$, which is large even for moderate values of K. Even for  $K=4$, we would need a large number of subjects and cells to properly estimate the 33 parameters. A solution would be to explore alternative model parameterizations that could be used to reduce the number of required parameters.
For example, one could assume that the hyper-parameters are constant across biomarker combinations, \textit{i.e.}, $\alpha^{\{s,u\}}_{k}=\alpha^{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
In reality, such a model would be unrealistic given that certain stimulations are known to induce expression of certain biomarkers more than others.
More exploratory work will need to be done in this area once high dimensional single-cell level data with large number of samples become available.
